A thermodynamic view of dusty protoplanetary disks

### 2017-05-02 summary.tex
misc edits to summary

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extrema in the generalized
potential vorticity 
%\begin{align}
 $\mathcal{V}_\mathrm{g} =
 \kappa^2\mathcal{S}_\mathrm{g}^{-2/\gamma}/2\Omega\Sigma_\mathrm{g}$, 
%\end{align} 
where $\mathcal{S}_\mathrm{g} = P/\Sigma_\mathrm{g}^\gamma$. Here $P$
should be interpreted as the vertically integrated pressure. The
non-linear result of the RWI is vortex formation \citep{li01}.

The dusty/adiabatic gas analogy imply the condition for RWI in a
polytropic dusty gas is given by extrema in  
\begin{align}
  \mathcal{V} =
  \frac{\kappa^2}{2\Omega\Sigma}\frac{1}{K^{2/\Gamma}}\left(1+
    \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}}\right)^2.    
\end{align}
Thus we expect that RWI may also be triggered by extrema 
in the dust-to-gas ratio, e.g. narrow dust rings/gaps. This may lead
to direct formation of dusty vortices, as opposed to dust-trapping by  
a pre-existing gas vortex \citep{barge95,lyra13}.

@@

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear instability discovered in pure gas disks
\citep{marcus15,umurhan16d}. It operates by having finite-amplitude perturbations
exciting `critical layers' which role 
up into vortices, exciting further critical layers. The process
repeats itself and the result is an array of vortices. These critical 
layers are locations where the intrinsic wave frequency of a 
disturbance matches the local vertical buoyancy frequency 
\citep{marcus13}.

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas ($\tcool\OmK\gg
1$). This limits its applicability to typical protoplanetary disks at
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk.

On the other hand, we have show that dust-loading an isothermal disk
also induces an effective buoyancy. This effect is maximized when dust is
perfectly coupled to the gas. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce a `dusty ZVI'. This may allow the ZVI
to operate at $\gtrsim 1$AU.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not add any new 
equations to evolve the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal
conduction, which is also common in modern hydrodynamics codes.

@@

\subsection{Higher order corrections}
{\bf notes for self. maybe exclude in final. }

The one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\widetilde{\bm{v}}|^2)$ after converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations

@@

In a following paper we will include $\bm{T}$ in the one-fluid
equations and futher explore the analogy between dusty-gas and viscous
hydrodynamics.

@@

\section{Summary}\label{summary}
In this paper we develop a physical analogy between isothermal 
dusty-gas and adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite dust-gas drag, $\tstop\neq 0$, the dust content of a 
parcel of the mixture is no longer conserved. The parcel 
can exchange dust particles with neighboring parcels. % owing 
%to dust-gas relative drift. 
This is analogous to heat exchange between a parcel of pure gas and its surroundings.

We explicitly show that for a fixed gas equation of state, the  
evolutionary equation for $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
Finite dust-gas friction then appears as an energy
source term.  This analogy with standard 
hydrodynamics with cooling/heating allow us to obtain dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

We obtain the Solberg-Hoiland criteria for axisymmetric stability for
strictly isothermal, perfectly-coupled dusty gas.  
Applying this to typical protoplanetary disks, we find that 
the vertical shear associated with dust 
 layers \emph{cannot} lead to axisymmetric  
  instabilities, however thin the dust layer is.   
%This is consistent with the fact that only 
%\emph{non-axisymmetric} Kelvin-Helmholtz instabilities have been 
%observed in numerical simulations in the perfectly-coupled limit 
%\citep{chiang08,barranco09,lee10}. 
Instead, sharp radial  edges in the dust-to-gas ratio could destabilized the
disk, as these imply sharp gradients in the disk's effective entropy
profile. % {\bf presumably this limits how sharp dust rings can be}

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show that the SI occurs because the evolution of gas
pressure lags behind  dust density; a general property of 
overstabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work, which goes into amplifying
oscillation amplitudes. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

>>>>

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extremum in the generalized
potential vorticity 
%\begin{align}
 $\mathcal{V}_\mathrm{g} =
 \kappa^2\mathcal{S}_\mathrm{g}^{-2/\gamma}/2\Omega\Sigma_\mathrm{g}$, 
%\end{align} 
where $\mathcal{S}_\mathrm{g} = P/\Sigma_\mathrm{g}^\gamma$. Here $P$
should be interpreted as the vertically integrated pressure. The
non-linear result of the RWI is vortex formation \citep{li01}.

The dusty/adiabatic gas analogy imply the condition for RWI in a
polytropic dusty gas disk is an extremum   
\begin{align}
  \mathcal{V} =
  \frac{\kappa^2}{2\Omega\Sigma}\frac{1}{K^{2/\Gamma}}\left(1+
    \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}}\right)^2.    
\end{align}
Thus we expect that RWI may also be triggered by extrema 
in the dust-to-gas ratio, e.g. narrow dust rings/gaps. This may lead
to direct formation of dusty vortices, as opposed to dust-trapping by  
a pre-existing gas vortex \citep{barge95,lyra13}.

@@

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear instability discovered in pure gas disks
\citep{marcus15,umurhan16d}. It operates by having finite-amplitude perturbations
exciting `critical layers' which roll 
up into vortices, exciting further critical layers. The process
repeats itself and the result is an array of vortices. These critical 
layers are locations where the intrinsic wave frequency of a 
disturbance matches the local vertical buoyancy frequency 
\citep{marcus13}.

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas ($\tcool\OmK\gg
1$). This limits its applicability in typical protoplanetary disks to 
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk.

On the other hand, we have show that dust-loading an isothermal disk
also induces an effective buoyancy. % This effect is maximized when dust is
% perfectly coupled to the gas
. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce a `dusty ZVI'. This may allow the ZVI
to operate at $\gtrsim 1$AU in PPDs.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not implement a separate 
equation for the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ (or $K$) is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal
conduction, which is also common in modern hydrodynamics codes.

@@

\subsection{Higher order corrections}
{\bf notes for self. maybe exclude in final. maybe move into appendix}

The simplified one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\widetilde{\bm{v}}|^2)$ after converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations

@@

% In a following paper we will include $\bm{T}$ in the one-fluid
% equations and futher explore the analogy between dusty-gas and viscous
% hydrodynamics.

@@

\section{Summary}\label{summary}
In this paper we develop the analogy between isothermal 
dusty-gas and adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite drag, $\tstop\neq 0$, the dust content of a 
parcel of the dusty-gas mixture is no longer conserved. The parcel 
can exchange dust particles with neighboring parcels. % owing 
%to dust-gas relative drift. 
This is analogous to heat exchange between a parcel of pure gas and
its surroundings.

We explicitly show that for a fixed gas equation of state, the  
evolutionary equation for the dust-to-gas ratio $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
The effect of finite dust-gas friction appears as an energy
source term.  This analogy with standard 
hydrodynamics with cooling/heating allow us to find dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

We obtain an analog of the Solberg-Hoiland criteria for
axisymmetric stability for strictly isothermal, perfectly-coupled dusty gas.  
Applying this to typical protoplanetary disks, we find that 
the vertical shear associated with dust 
 layers \emph{cannot} lead to axisymmetric  
  instabilities, however thin the dust layer is.    
Instead, sharp radial  edges in the dust-to-gas ratio could destabilized the
disk, as these imply sharp gradients in the disk's effective entropy
profile. Alternatively, if the dust is vertically well-mixed, then any
radial gradient in $\rhod/\rhog$ can destabilize the disk.

 % {\bf presumably this limits how sharp dust rings can be}

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show in SI the evolution of gas
pressure lags behind dust density; and is in fact a general property of 
overstabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the dusty-gas mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work that amplifies 
oscillations. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.
